---
title: "HEATMAP"
author: "by Steph and Ale"
output:
  html_document:
    theme: readable
    highlight: haddock
    toc: yes
    toc_depth: 2
    toc_float: yes
bibliography: references.bib
---

# 

This script is to generate, explore and visualize a heatmap of relative abundance with transformed data using 'pheatmap' package [@kolde2015package].

In this tutorial , Let's use the same test data used in the previous script (PCA)[^1]. For this purpose Let's load the data and Libraries that it is will be used.

[^1]: Data provided from Ph D Alejandra Miranda of her experiment (real data)

# Libraries

```{r Libraries, message=FALSE, warning=FALSE, include=TRUE}
library(ALDEx2)
library(CoDaSeq)
library(tibble)
library (dplyr)
library(pheatmap)
library(propr)
library(vegan)
library(tidyr)


```

# Data

\
1. Load data from a tsv file using *read.delim*\
2. Rename data to a shorter name\
3. Visualizing data (only a portion)\

```{r, warning=FALSE}
soil_species <- read.delim("../Data/soil_species_bacteria.txt", 
                           row.names=1, check.names = FALSE)
soil <- soil_species
head(soil[,1:3], 5)

metadata <- read.delim("../Data/soil_maize_metadata.txt", check.names = FALSE,
                       header = TRUE, row.names = 1)
head(metadata, 5)
metadata <- data.frame(t(metadata), check.names = FALSE)
head (metadata[,1:3], 5)

#Make a column with the sum of every sample
soil$sum <- rowSums(soil)
tail(rev(soil)[1])
soil <- soil[order(-soil$sum),]
head(rev(soil)[1])
soil <- soil[-37]
head(rev(soil)[1])
soil <- soil[1:50,] 

soil <- data.frame(t(soil))
metadata <- data.frame(t(metadata))

data <- merge(metadata, soil, by=0, all=TRUE)

##CONCATENATE VARIABLES (SUBSTRAUM/PERIOD)

data <- unite(data, till_nit,c(4,5), sep=";")
data <- data[,c(-1,-2,-3,-5)]

data <- data %>% group_by(till_nit) %>% summarise_all(funs(sum))
row.names(data) <- data$till_nit
data <- data.frame(data)
data <- data [-1]
head(data[,1:3], 2)
data <- data.frame(t(data))

```

# Data Transformation

```{r}
# Compostinal analyses ----------------------------------------------------
#Generate the aldex object for Expectued values (variables)
conds <- c(rep("cond1", 2), rep("cond2", 2))

#Estimate the distribution of CLR values
data.x <- aldex.clr(data, conds, mc.samples=256, verbose=FALSE)

##Estimate expected values of effect sizes
data.e <- aldex.effect(data.x, include.sample.summary = TRUE, verbose = FALSE)

#use for ordination and dendogram
E.E.clr <- t(data.e[,grep("rab.sample", colnames(data.e))])
rownames(E.E.clr) <- gsub("rab.sample.", "", rownames(E.E.clr))
exp <- apply(E.E.clr,1,function(x)2^x)
E.clr <- t(apply(exp,2,function(x)log2(x)-mean(log2(x))))

E.clr <- data.frame(t(E.clr))

head(E.clr[,1:2], 5)

```

# Setting ranks and Visualizing

```{r}
# Heatmap -----------------------------------------------------------------

min(E.clr)   ###Para ver cantidad minima y tomar en cuenta para mutar valores
max(E.clr)   ###Cantidad maxima

E.clr_heat <-E.clr %>% rownames_to_column( var = "treatments")
head(E.clr_heat[,1:2])

#Cambiar valores de columnas que tienen "_" por los asignados
#(se puede poner cualquier simbolo que coincida con los nombres de las col)

E.clr_heat2 <-E.clr_heat%>% mutate_at(vars(contains(".")), funs(R = case_when(
    . > -1.5 & .  <= -1.0 ~ 0,
  . >  -1.0 & .  <= -0.5 ~ 1,
  . >  -0.5 & .  <= 0 ~ 2,
  . >  0 & .  <= 0.5 ~ 3,
  . >  0.5 & .  <= 1 ~ 4,
  . >  1 & .  <= 1.5 ~ 5,
  . >  1.5 & .  <= 2 ~ 6,
  . >  2 & .  <= 2.5 ~ 7,
  . >  2.5 ~ 8)))

#Regresar la primer columna, que contiene los phyla, como nombre de filas y no como columna
row.names(E.clr_heat2) <-E.clr_heat2$treatments

#Remove columns without mutations
E.clr_heat3 <- E.clr_heat2[-1:-5]

#Add row with scsoil
colnames(E.clr_heat3)

E.clr_heat3_esc <- add_row(E.clr_heat3, No.No_R = 0, No.Yes_R = 4, Yes.Yes_R = 8, .before = 1,)


#File with color palette for heatmap
my_palette <- colorRampPalette(c("white", "yellow", "red"))(n=599)
ale_colors <- colorRampPalette(c("white", "lightgoldenrod1", "darksalmon" ,"coral2", "maroon4"))(n=599)

#Heatmap
pheatmap(E.clr_heat3_esc, color = ale_colors, main="Heat map", cluster_cols = T, cluster_rows = F, fontsize = 7, clustering_distance_cols = "euclidean")
 

```

## **References**

::: {#refs}
:::
